---
title: "Customized_upset_plot"
author: "Chenxin Li"
date: "1/22/2020"
output: 
  html_notebook:
    number_sections: yes
    toc: yes
    toc_float: yes
  html_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Dependencies 
If you have Tidyverse and cowplot you are mostly covered. If you want to save plot as .svg you will need the R package svglite. If you are using Mac, you need to install XQuart. Link = https://www.xquartz.org/ 
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(readr)
library(readxl)
library(stringr)
library(svglite)  
library(cowplot)
```

#data
```{r}
#this chunck generates the data 
#Load your data here when you do your acutal thing 
#e.g. data1 <- read_delim(...) 

set.seed(233333) #set seed to make simulations reproducible, not needed for actual data 

data1 <- data.frame(
  "elements" = round(runif(n = 30, min = 0, max = 100))
) %>% 
  distinct(elements, .keep_all = T)

data2 <- data.frame(
  "elements" = round(runif(n = 150, min = 0, max = 200))
) %>% 
  distinct(elements, .keep_all = T)

data3 <- data.frame(
  "elements" = round(runif(n = 200, min = 0, max = 300))
) %>% 
  distinct(elements, .keep_all = T)

data4 <- data.frame(
  "elements" = round(runif(n = 60, min = 0, max = 100))
) %>% 
  distinct(elements, .keep_all = T)
```

#find specifics 
You can write a loop or something (apply or something in purr) to achieve all these 
```{r}
find_specific <- function(set1, against){
  set1 %>% 
    filter(elements %in% against$elements == F) #In your data, the col name won't be "element". Change it here. 
}
```

```{r}
data1_specific <- data1 %>% 
  find_specific(against = rbind(data2, data3, data4))

data2_specific <- data2 %>% 
  find_specific(against = rbind(data1, data3, data4))

data3_specific <- data3 %>% 
  find_specific(against = rbind(data1, data2, data4))

data4_specific <- data4 %>% 
  find_specific(against = rbind(data1, data2, data3))
```

```{r}
ndata1_spec <- data1_specific %>% nrow()
ndata2_spec <- data2_specific %>% nrow()
ndata3_spec <- data3_specific %>% nrow()
ndata4_spec <- data4_specific %>% nrow()

#the above 2 chunks can be simplied by adding %>% nrow() to the find_specifc function. 
#However, if you want examine the elements in each intersections, I'd recommend take nrow() later 
```


#find overlaps 
```{r}
find_2way <- function(set1, set2){
  set1 %>% 
    filter(set1$elements %in% set2$elements)  #In your data, the col name won't be "element". Change it here. 
  
} 
```

```{r}
data1by2 <- find_2way(data1, data2)
data1by3 <- find_2way(data1, data3)
data1by4 <- find_2way(data1, data4)

data2by3 <- find_2way(data2, data3)
data2by4 <- find_2way(data2, data4)

data3by4 <- find_2way(data3, data4)
```

```{r}
ndata1by2 <- nrow(data1by2)
ndata1by3 <- nrow(data1by3)
ndata1by4 <- nrow(data1by4)

ndata2by3 <- nrow(data2by3)
ndata2by4 <- nrow(data2by4)

ndata3by4 <- nrow(data3by4)
```

```{r}
find_3way <- function(set1, set2, set3){
  set1 %>% 
    filter(set1$elements %in% set2$elements) %>%  #In your data, the col name won't be "element". Change it here. 
    filter(elements %in% set3$elements) 
}
```

```{r}
data1by2by3 <- find_3way(data1, data2, data3)
data2by3by4 <- find_3way(data2, data3, data4)
data3by4by1 <- find_3way(data3, data4, data1)
data4by1by2 <- find_3way(data4, data1, data2)
```

```{r}
ndata1by2by3 <- data1by2by3 %>% nrow()
ndata2by3by4 <- data2by3by4 %>% nrow()
ndata3by4by1 <- data3by4by1 %>% nrow()
ndata4by1by2 <- data4by1by2 %>% nrow()
```


```{r}
#the highest order of overlap in this example is 4-way intersection, 
#which is commonly the largest in most of the Venn Diagrams I've seen. 
#You can go further beyond w/ upset plots if you like. 

data1by2by3by4 <- data1 %>% 
  filter(elements %in% data2$elements) %>% 
  filter(elements %in% data3$elements) %>% 
  filter(elements %in% data4$elements)

ndata1by2by3by4 <- data1by2by3by4 %>% nrow()
```

#total set size
```{r}
total <- data.frame(
  "sets" = c("data1", "data2", "data3", "data4"),
  "sizes" = c(nrow(data1), nrow(data2), nrow(data3), nrow(data4))
)

total #this will be the basis for the upperleft bar plot 
```


```{r}
upperleft <- total %>% 
  ggplot(aes(x = sets, y= sizes)) +
  geom_hline(yintercept = -Inf, size = 1.5) +
  geom_vline(xintercept = -Inf, size = 1.5) +
  geom_bar(stat = "identity", aes(fill = sets), alpha = 0.8) +
  geom_text(aes(label = as.character(sizes)), size = 5, angle = 90, hjust = 0, y = 1, fontface = "bold") +
  scale_fill_manual(values = c("orangered3", "tomato1", "seagreen", "dodgerblue2"), #this is my own custom palette  
                     limits = c("data1", "data2", "data3", "data4")) +              #feel free to use something else     
  scale_x_discrete(labels = NULL) +
  scale_y_continuous(labels = NULL) +
  labs(x = NULL,
       y = "set size") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(text = element_text(size= 18, face="bold")) +
  theme(axis.text.x=element_text(colour = "black", angle = 45, hjust = 1)) +
  theme(axis.text.y=element_text(colour = "black")) +
  theme(panel.grid = element_blank()) 

upperleft
```

#overlap matrix
```{r}
sets <- c("1", "2", "3", "4")
category <- c(
    "data1_specific",
    "data2_specific",
    "data3_specific",
    "data4_specific",
    "data1by2", 
    "data1by3",
    "data1by4",
    "data2by3",
    "data2by4",
    "data3by4",
    "data1by2by3",
    "data2by3by4",
    "data3by4by1",
    "data4by1by2",
    "data1by2by3by4" 
    )

overlap_matrix <- expand.grid(  #generate all the combinations 
  "sets" = sets,
  "category" = category
) %>% 
  as.data.frame()  %>%    #determine the intersection: a character col of Y or N. 
  mutate(intersect = case_when(
    str_detect(category, sets %>% as.character()) ~ "Y",
    T ~ "N"
  ))  

overlap_matrix
#this will be the bais for lower left plot 
```

```{r}
lowerleft <- overlap_matrix %>% 
  mutate(category = factor(rev(category))) %>% #in the colored matrix the first y value apprears in the bottom, 
                                               #so the order need to be reversed 
  ggplot(aes(x = sets, y = category))+
  geom_tile(aes(fill = sets, alpha = intersect), color = "black", size = 1.5) +
  scale_fill_manual(values = c("orangered1", "tomato1", "seagreen", "dodgerblue2"),
                    limits = c("1", "2", "3", "4")) +
  scale_alpha_manual(values = c(0.8, 0),  #color the grid for Y, don't color for N. 
                     limits = c("Y", "N")) +
  scale_y_discrete(labels = NULL) +
  scale_x_discrete(labels = NULL) +
  labs(x = " ", #I left a white space here so that it will be better aligned w/ the bar plot next to it 
       y = "overlap") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(text = element_text(size= 18, face="bold")) +
  theme(axis.text.x=element_text(colour = "black")) +
  theme(axis.text.y=element_text(colour = "black")) +
  theme(panel.grid = element_blank())

lowerleft
```

#produce the legend 
the legend will show up at upper right 
```{r}
upperright <- get_legend(
  total %>% 
  ggplot(aes(x = sets, y= sizes)) +
  geom_hline(yintercept = -Inf, size = 1.5) +
  geom_vline(xintercept = -Inf, size = 1.5) +
  geom_bar(stat = "identity", aes(fill = sets), alpha = 0.8) +
  geom_text(aes(label = as.character(sizes)), size = 5, angle = 90, hjust = 0, y = 1, fontface = "bold") +
  scale_fill_manual(values = c("orangered3", "tomato1", "seagreen", "dodgerblue2"), #this is my own custom palette  
                     limits = c("data1", "data2", "data3", "data4")) +              #feel free to use something else     
  scale_x_discrete(labels = NULL) +
  scale_y_continuous(labels = NULL) +
  labs(x = NULL,
       y = "set size") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(text = element_text(size= 18, face="bold")) +
  theme(axis.text.x=element_text(colour = "black", angle = 45, hjust = 1)) +
  theme(axis.text.y=element_text(colour = "black")) +
  theme(panel.grid = element_blank()) 
)

#same code as upper left, but w/ legend.position = right to get the legend   
plot_grid(upperright)
```

#lowerright  
```{r}
overlap_sizes <- data.frame(
  "overlap_sizes" = c(ndata1_spec, ndata2_spec, ndata3_spec, ndata4_spec, 
                      ndata1by2, ndata1by3, ndata1by4, ndata2by3, ndata2by4, ndata3by4, 
                      ndata1by2by3, ndata2by3by4, ndata3by4by1, ndata4by1by2,
                      ndata1by2by3by4),
  "category" = category 
) %>% 
  mutate(category = factor(category, levels = rev(category)))   #again order needs to be reversed 

overlap_sizes #this will be the basis for bar plot at the lower right 
```

```{r}
lowerright <- overlap_sizes %>% 
  ggplot(aes(x = category, y = overlap_sizes)) +
  geom_hline(yintercept = -Inf, size = 1.5) +
  geom_vline(xintercept = -Inf, size = 1.5) +
  geom_bar(stat = "identity", fill = "grey80", color = NA, alpha = 0.8) +
  geom_text(aes(label = overlap_sizes, y = 0), size = 5, hjust = 0, vjust = 0.5, fontface = "bold") +
  scale_y_continuous(labels = NULL) +
  scale_x_discrete(labels = NULL) +
  labs(y = "intersect. sizes",
       x = NULL) +
  theme_minimal() +
  theme(text = element_text(size= 18, face="bold")) +
  theme(axis.text.x=element_text(colour = "black", angle = 45, hjust = 1)) +
  theme(axis.text.y=element_text(colour = "black")) +
  theme(panel.grid = element_blank()) + 
  coord_flip()

lowerright
```


#put them all together
```{r}
plot_grid(upperleft, upperright, lowerleft, lowerright, 
          nrow = 2, 
          ncol = 2,
          rel_heights = c(1, 4), #the more rows in the lower part, the longer it should be
          rel_widths = c(1, 0.8))

ggsave("upset_full.svg", height = 7, width = 3.5) #this should be a tall & skinny plot 
#I prefer .svg, but you can also save as phd or png 
#I will open up the .svg file and mannually adjust the size until it's good
#check that nothing is cut off from the plot 
#png is for twitter posting 
ggsave("upset_full.png", height = 7, width = 3.5)
```

#subsetting the inserctions
Say you are only interested in the overlaps w/ data3 

```{r}
lowerleft_3only <- overlap_matrix %>% 
  filter(str_detect(category, "by3|data3")) %>% #subset out things that overlaps w/ data3 
  mutate(category = factor(rev(category))) %>%
  ggplot(aes(x = sets, y = category))+
  geom_tile(aes(fill = sets, alpha = intersect), color = "black", size = 1.5) +
  scale_fill_manual(values = c("orangered1", "tomato1", "seagreen", "dodgerblue2"),
                    limits = c("1", "2", "3", "4")) +
  scale_alpha_manual(values = c(0.8, 0),
                     limits = c("Y", "N")) +
  scale_y_discrete(labels = NULL) +
  scale_x_discrete(labels = NULL) +
  labs(x = " ", #I left a white space here so that it will be better aligned w/ the bar plot next to it 
       y = "overlap") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(text = element_text(size= 18, face="bold")) +
  theme(axis.text.x=element_text(colour = "black")) +
  theme(axis.text.y=element_text(colour = "black")) +
  theme(panel.grid = element_blank())

lowerleft_3only
```
```{r}
lowerright_3only <- overlap_sizes %>% 
  filter(str_detect(category, "by3|data3")) %>% #subset out things that overlaps w/ data3 
  ggplot(aes(x = category, y = overlap_sizes)) +
  geom_hline(yintercept = -Inf, size = 1.5) +
  geom_vline(xintercept = -Inf, size = 1.5) +
  geom_bar(stat = "identity", fill = "grey80", color = NA, alpha = 0.8) +
  geom_text(aes(label = overlap_sizes, y = 0), size = 5, hjust = 0, vjust = 0.5, fontface = "bold") +
  scale_y_continuous(labels = NULL) +
  scale_x_discrete(labels = NULL) +
  labs(y = "intersect. sizes",
       x = NULL) +
  theme_minimal() +
  theme(text = element_text(size= 18, face="bold")) +
  theme(axis.text.x=element_text(colour = "black", angle = 45, hjust = 1)) +
  theme(axis.text.y=element_text(colour = "black")) +
  theme(panel.grid = element_blank()) + 
  coord_flip()

lowerright_3only
```
```{r}
plot_grid(upperleft, upperright, lowerleft_3only, lowerright_3only, 
          nrow = 2, 
          ncol = 2,
          rel_heights = c(1, 2.5), #this time the lower part can be shorter because there're less rows 
          rel_widths = c(1, 0.8))

ggsave("upset_3only.svg", height = 4.5, width = 3.5) #this time the plot can be shorter 
#the best aspect ratio is when the grids in the colored matrix are perfect squares 
ggsave("upset_3only.png", height = 4.5, width = 3.5) 
```


#extensions
the right side of the upset plot can be extended 
This works much better when you don't have any empty sets 
Actually this ONLY works well when you don't have any empty sets 
```{r}
#Do you have any empty sets? 
overlap_sizes %>% filter(overlap_sizes == 0)
```

```{r}
full_data <- rbind(data1, data2, data3, data4)
#simulations are for examples only. No need this for actual data 
#just giving a value to each elements
#your data will already have this 

element_values <- data.frame(
  "elements" = 1:300, 
  "values" = rnorm(mean = 50, sd = 10, n = 300)
)

head(element_values)
```

```{r}
intersection_data <- rbind(
  data1_specific %>% 
    mutate(category = "data1_specific"), 
  data2_specific %>% 
    mutate(category = "data2_specific"), 
  data3_specific %>% 
    mutate(category = "data3_specific"), 
  data4_specific %>% 
    mutate(category = "data4_specific"), 
  data1by2 %>% 
    mutate(category = "data1by2"),
  data1by3 %>% 
    mutate(category = "data1by3"),
  data1by4 %>% 
    mutate(category = "data1by4"),
  data2by3 %>% 
    mutate(category = "data2by3"),
  data2by4 %>% 
    mutate(category = "data2by4"),
  data3by4 %>% 
    mutate(category = "data3by4"),
  data1by2by3 %>% 
    mutate(category = "data1by2by3"),
  data2by3by4 %>% 
    mutate(category = "data2by3by4"),
  data3by4by1 %>% 
    mutate(category = "data3by4by1"),
  data4by1by2 %>% 
    mutate(category = "data4by1by2"),
  data1by2by3by4 %>% 
    mutate(category = "data1by2by3by4")
) %>% 
  inner_join(element_values, by = "elements") %>% 
  rbind(overlap_sizes %>% 
    filter(overlap_sizes == 0) %>% 
    mutate(elements = NA) %>% 
    mutate(values = NA) %>% 
    select(elements, category, values)) %>%    #this is the workaround for empty sets)
  mutate(category = factor(category, levels = rev(c(
    "data1_specific","data2_specific", "data3_specific", "data4_specific",
    "data1by2", "data1by3", "data1by4",
    "data2by3", "data2by4", "data3by4",
    "data1by2by3", "data2by3by4", "data3by4by1", "data4by1by2",
    "data1by2by3by4"
  )))) 

head(intersection_data) 
```
```{r}
library(ggbeeswarm) #this is required for quasirandom jittering of dots 
#if you don't care you can just use geom_jitter 
```

```{r}
farright_plot <- intersection_data %>% 
  filter(is.na(category) == F) %>% 
  ggplot(aes(x = category, y = elements)) +
  geom_hline(yintercept = -Inf, size = 1.5) +
  geom_vline(xintercept = -Inf, size = 1.5) +
  geom_quasirandom(shape = 21, color = "white", fill = "grey20", position = position_jitter(0.1), alpha = 0.8) +
  geom_boxplot(outlier.shape  = NA, width = 0.7, fill = "skyblue1", alpha = 0.5) +
  theme_minimal()+
  scale_y_continuous(labels = NULL) +
  scale_x_discrete(labels = NULL) +
  labs(y = "response",
       x = NULL) +
  theme_minimal() +
  theme(text = element_text(size= 18, face="bold")) +
  theme(axis.text.x=element_text(colour = "black", angle = 45, hjust = 1)) +
  theme(axis.text.y=element_text(colour = "black")) +
  coord_flip()

farright_plot

#in your actual data you can color the boxes or dots w/ factors of interest 
```
```{r}
#If you have empty sets the alignment won't be correct 

plot_grid(upperleft, upperright, NULL, 
          lowerleft, lowerright, farright_plot,  
          nrow = 2, 
          ncol = 3,
          rel_heights = c(1, 3),  
          rel_widths = c(0.8, 1, 1)) #the more stuff on the right, the thinner the left side has to be. 

ggsave("upset_extended.svg", height = 7, width = 6) #this time the plot needs to be wider
ggsave("upset_extended.png", height = 7, width = 6) 
```




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
