---
title: "Quick_start.Rmd"
author: "Chenxin Li"
date: "2023-01-20"
output: html_notebook 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This is a workflow for set/intersect visualization using [UpSet plots](https://github.com/hms-dbmi/UpSetR). 
The upstream segment of the workflow (intersect size determination) is based on the re-implementation of `UpSet` plots by the package [ComplexHeatmap](https://jokergoo.github.io/ComplexHeatmap-reference/book/). 
Basic list, data frame, and plot handling was provided by the [tidyverse](https://www.tidyverse.org/). 
Lastly, construction of composite plots is provided by [patchwork](https://cran.r-project.org/web/packages/patchwork/vignettes/patchwork.html).

In traditional [upset plots](https://upset.app/), intersects/subsets are indicated by dots. 
When two dots are connected by a line, it represents the distinct intersect between the two sets. 
Set and intersect sizes are then represented by bars. 

The workflow produces customized upset plot where intersects/subsets are indicated by a heatmap. 
![insert quick start here]() 

In the heatmap (lower left panel), each column is a set, and each row is a intersect/subset. 
The color in a tile represents elements in the intersect/subset are present in a set. 
As in traditional upset plots, set and intersect sizes are then represented by bars. 
In addition, upset plots can be extended. 
Mean separation plots (e.g., box plot, bar plot) and annotations (heatmaps) can be added to the sides of the upset plot using `patchwork`.
![insert full extended here]()

Finally, the intersect/subsets can be filtered. 
![insert set 3 example here]() 

# Dependencies 
```{r}
library(tidyverse) 
library(patchwork) 
library(ComplexHeatmap)

library(RVenn) # Only required if you want Venn diagrams 
library(RColorBrewer) # This is for the colors only, not actually necessary
```

Auxiliary dependencies 

* For 2-3 sets, Venn diagrams can be made readily using the [RVenn package](https://cran.r-project.org/web/packages/RVenn/vignettes/vignette.html). The `ggVenn()` function from `RVenn` produces a ggplot object that is a Venn Diagram. 
* The official way to install `ComplexHeatmap` is via `devtools::install_github("jokergoo/ComplexHeatmap")`, which requires the `devtools` package.
* For mean separation plots, a suggested package is [ggbeeswarm](https://github.com/eclarke/ggbeeswarm), a violin plot, but with actual data points. 
* For color palettes, suggested are `viridis` and `RColorBrewer` packages. 
* If you want to save plot as .svg file, you may need the R package `svglite`. If you are using Mac, you may need to install XQuart. [Link](https://www.xquartz.org/)

# Getting started
Here are example scripts for 3 sets. 
The workflow is scalable to more sets, as intersect size calculation is automatic (provided by `ComplexHeatmap`). 
However, for a large number of sets, number of subsets increases geometrically, and thus filtering for subset of interest will be important. 

# Data 
```{r}
my_list <- list(
  data1 = letters[1:10], 
  data2 = letters[3:13], 
  data3 = letters[6:18])
```

# If you want a Venn diagram
```{r}
my_object <- RVenn::Venn(my_list)

ggvenn(
  my_object, slice = 1:3, 
  thickness = 0.5,
  alpha = 0.5, 
  fill = brewer.pal(8, "Set2")
) +
  theme_void() +
  theme(
    legend.position = "none"
  )

ggsave("../Results/VennDiagram_quick_start.svg", height = 4, width = 4, bg = "white")
ggsave("../Results/VennDiagram_quick_start.png", height = 4, width = 4, bg = "white")
```
# ComplexHeatmap for heavy-lyfting 
```{r}
comb_mat <- make_comb_mat(my_list)
my_names <- set_name(comb_mat)
my_names
```

## Total set size 
```{r}
my_set_sizes <- set_size(comb_mat) %>% 
  as.data.frame() %>% 
  rename(sizes = ".") %>% 
  mutate(Set = row.names(.)) 

p1 <- my_set_sizes %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  ggplot(aes(x = Set, y= sizes)) +
  geom_bar(stat = "identity", aes(fill = Set), alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sizes), 
            size = 5, angle = 90, hjust = 0, y = 1) +
  scale_fill_manual(values = brewer.pal(4, "Set2"),  # feel free to use some other colors  
                     limits = my_names) + 
  labs(x = NULL,
       y = "Set size",
       fill = NULL) +
  theme_classic() +
  theme(legend.position = "right",
        text = element_text(size= 14),
        axis.ticks.y = element_blank(),
        axis.text = element_blank()
        ) 

p1
```
## Legend 
It's not easy to extract legend.
But we can write a function for that.  
```{r}
get_legend <- function(p) {
  tmp <- ggplot_gtable(ggplot_build(p))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  legend
}
```

```{r}
p2 <- get_legend(p1)
```

## Overlap sizes
```{r}
my_overlap_sizes <- comb_size(comb_mat) %>% 
  as.data.frame() %>% 
  rename(overlap_sizes = ".") %>% 
  mutate(category = row.names(.))

p3 <- my_overlap_sizes %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>% 
  ggplot(aes(x = category, y = overlap_sizes)) +
  geom_bar(stat = "identity", fill = "grey80", color = NA, alpha = 0.8, width = 0.7) +
  geom_text(aes(label = overlap_sizes, y = 0), 
            size = 5, hjust = 0, vjust = 0.5) +
  labs(y = "Intersect sizes",
       x = NULL) +
  theme_classic() +
  theme(text = element_text(size= 14, color = "black"),
        axis.text =element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(hjust = 0),
        ) +
  coord_flip()

p3
```

## Overlap matrix 
```{r}
my_overlap_matrix <- str_split(string = my_overlap_sizes$category, pattern = "", simplify = T) %>% 
  as.data.frame() 

colnames(my_overlap_matrix) <- my_names

my_overlap_matrix_tidy <- my_overlap_matrix %>% 
  cbind(category = my_overlap_sizes$category) %>% 
  pivot_longer(cols = !category, names_to = "Set", values_to = "value") %>% 
  full_join(my_overlap_sizes, by = "category") %>% 
  full_join(my_set_sizes, by = "Set")

p4 <- my_overlap_matrix_tidy %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>%  
  mutate(Set = reorder(Set, sizes)) %>%  
  ggplot(aes(x = Set, y = category))+
  geom_tile(aes(fill = Set, alpha = value), color = "grey30", size = 1) +
  scale_fill_manual(values = brewer.pal(4, "Set2"), # feel free to use other colors 
                    limits = my_names) +
  scale_alpha_manual(values = c(0.8, 0),  # color the grid for 1, don't color for 0. 
                     limits = c("1", "0")) +
  labs(x = "Sets",  
       y = "Overlap") +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(color = "black", size= 14),
        panel.grid = element_blank(),
        axis.text = element_blank()
        )

p4
```
# Put them together 
```{r}
wrap_plots(p1, p2, p4, p3, 
          nrow = 2, 
          ncol = 2,
          heights = c(1, 2), # the more rows in the lower part, the longer it should be
          widths = c(1, 0.8),
          guides = "collect") &
  theme(legend.position = "none")

ggsave("../Results/quick_start.svg", height = 3.5, width = 3, bg = "white") 
# this should be a tall & skinny plot 
# I prefer .svg, but you can also save as phd or png 
# I will open up the .svg file and mannually adjust the size until it's good
# check that nothing is cut off from the plot 
# png is for twitter posting 
ggsave("../Results/quick_start.png", height = 3.5, width = 3, bg = "white")
```

