---
title: "Customized_upset_plot"
author: "Chenxin Li"
date: "01/19/2023"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
This is a workflow to generate upset plots with custom aesthetics. 
This workflow uses real data. 
Data from: [Li et al., 2020](https://genome.cshlp.org/content/30/2/173.short) 

# Dependencies 
Main packages doing heavy lifting: 

* tidyverse
* ComplexHeatmap
* patchwork 

If you want to save plot as .svg you may need the R package svglite.
If you are using Mac, you may need to install XQuart. [Link](https://www.xquartz.org/)  

```{r}
library(tidyverse) 
library(patchwork) 
library(ComplexHeatmap)

library(RVenn) # Only required if you want Venn diagrams 
library(RColorBrewer) # This is for the colors only, not actually necessary  
library(ggbeeswarm) # Only required if you want quasirandom jittering of dots for mean separation plots 
```


# Data
```{r}
data1_full <- read_csv("../Data/data1.csv", col_types = cols())  # Ovary siRNA loci
data2_full <- read_csv("../Data/data2.csv", col_types = cols())  # Egg siRNA loci
data3_full <- read_csv("../Data/data3.csv", col_types = cols())  # Seedling siRNA loci
data4_full <- read_csv("../Data/data4.csv", col_types = cols()) # Sperm siRNA loci 
```

`col_types = cols()` argument supresses column type specification messages.

These are 4 different small RNA datasets. 

  * 1st column = genome location. This is what we’ll find intersection for.
  * 2nd column = their small RNA relative abundance (expression), in units of RPM.
  * last column = cell type.

we’ll ignore the rest of the columns.

Let’s rename the columns to match the code Or you can edit the code to match your colume names.
We’ll use the `$location` column  to find intersections first. We’ll worry about their expression in different cell types later.

```{r}
data1 <- data1_full %>% 
  mutate(elements = location) %>% 
  select(elements)

data2 <- data2_full %>% 
  mutate(elements = location) %>% 
  select(elements)

data3 <- data3_full %>% 
  mutate(elements = location) %>% 
  select(elements)

data4 <- data4_full %>% 
  mutate(elements = location) %>% 
  select(elements)
```


# Create set list object
```{r}
set_list <- list(ovary = data1$elements, 
                 egg = data2$elements, 
                 seedling = data3$elements, 
                 sperm = data4$elements) 

set_object <- RVenn::Venn(set_list)
```

The order of sets in the list matters. In this case: 

1. ovary (data1)
2. egg (data2)
3. seedling (data3)
4. sperm (data4) 

# Venn diagram for 2-3 sets 
```{r}
ggvenn(
  set_object, slice = 2:4, 
  thickness = 0.5,
  alpha = 0.5, 
  fill = c("tomato1", "seagreen", "dodgerblue2")
) +
  theme_void() +
  theme(
    legend.position = "none"
  )

ggsave("../Results/VennDiagram_real_data.svg", height = 6, width = 6, bg = "white")
ggsave("../Results/VennDiagram_real_data.png", height = 6, width = 6, bg = "white")
```

# ComplexHeatmap for heavy-lyfting 
```{r}
combination_matrix <- make_comb_mat(set_list)
set_names <- set_name(combination_matrix)
set_names
```

## Total set size 
```{r}
total_set_sizes <- set_size(combination_matrix) %>% 
  as.data.frame() %>% 
  rename(sizes = ".") %>% 
  mutate(Set = row.names(.)) 

upperleft <- total_set_sizes %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  ggplot(aes(x = Set, y= sizes)) +
  geom_bar(stat = "identity", aes(fill = Set), alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sizes), 
            size = 5, angle = 90, hjust = 0, y = 1) +
  scale_fill_manual(values = c("orangered3","tomato1", "seagreen", "dodgerblue2"),  # feel free to use some other colors  
                     limits = set_names) + 
  labs(x = NULL,
       y = "Set size",
       fill = NULL) +
  theme_classic() +
  theme(legend.position = "right",
        text = element_text(size= 14),
        axis.ticks.y = element_blank(),
        axis.text = element_blank()
        ) 

upperleft
```
## Legend 
It's not easy to extract legend.
But we can write a function for that.  
```{r}
get_legend <- function(p) {
  tmp <- ggplot_gtable(ggplot_build(p))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  legend
}
```

```{r}
upperright <- get_legend(upperleft)
```


## Overlap sizes
```{r}
overlap_sizes <- comb_size(combination_matrix) %>% 
  as.data.frame() %>% 
  rename(overlap_sizes = ".") %>% 
  mutate(category = row.names(.))

head(overlap_sizes)
```

Categories are ordered by frequency. 
```{r}
lowerright <- overlap_sizes %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>% 
  ggplot(aes(x = category, y = overlap_sizes)) +
  geom_bar(stat = "identity", fill = "grey80", color = NA, alpha = 0.8, width = 0.7) +
  geom_text(aes(label = overlap_sizes, y = 0), 
            size = 5, hjust = 0, vjust = 0.5) +
  labs(y = "Intersect sizes",
       x = NULL) +
  theme_classic() +
  theme(text = element_text(size= 14, color = "black"),
        axis.text =element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(hjust = 0),
        ) +
  coord_flip()

lowerright
```

## Overlap matrix 
```{r}
overlap_matrix <- str_split(string = overlap_sizes$category, pattern = "", simplify = T) %>% 
  as.data.frame() 

colnames(overlap_matrix) <- set_names

overlap_matrix_tidy <- overlap_matrix %>% 
  cbind(category = overlap_sizes$category) %>% 
  pivot_longer(cols = !category, names_to = "Set", values_to = "value") %>% 
  full_join(overlap_sizes, by = "category") %>% 
  full_join(total_set_sizes, by = "Set")

head(overlap_matrix_tidy)
```

```{r}
lowerleft <- overlap_matrix_tidy %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>%  
  mutate(Set = reorder(Set, sizes)) %>%  
  ggplot(aes(x = Set, y = category))+
  geom_tile(aes(fill = Set, alpha = value), color = "grey30", size = 1) +
  scale_fill_manual(values = c("orangered3","tomato1", "seagreen", "dodgerblue2"), # feel free to use other colors 
                    limits = set_names) +
  scale_alpha_manual(values = c(0.8, 0),  # color the grid for 1, don't color for 0. 
                     limits = c("1", "0")) +
  labs(x = "Sets",  
       y = "Overlap") +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(color = "black", size= 14),
        panel.grid = element_blank(),
        axis.text = element_blank()
        )

lowerleft
```
 
## Put them all together
```{r}
wrap_plots(upperleft, upperright, lowerleft, lowerright, 
          nrow = 2, 
          ncol = 2,
          heights = c(1, 4), # the more rows in the lower part, the longer it should be
          widths = c(1, 0.8),
          guides = "collect") &
  theme(legend.position = "none")

ggsave("../Results/real_data_upset_full.svg", height = 6, width = 3, bg = "white") 
# this should be a tall & skinny plot 
# I prefer .svg, but you can also save as phd or png 
# I will open up the .svg file and mannually adjust the size until it's good
# check that nothing is cut off from the plot 
# png is for twitter posting 
ggsave("../Results/real_data_upset_full.png", height = 6, width = 3, bg = "white")
```

# Subsetting the inserctions
## Using ComplexHeatmap 
Filtering of intersections can be done: 
Examples: 

* intersect size >= 4: `m2 <- combination_matrix[comb_size(combination_matrix) >= 4]`.
* degree == 3: `m3 <- combination_matrix[comb_degree(combination_matrix) == 3]`.
* only involving data1 & data2: `m4 <- combination_matrix[c("data1", "data2"), ]`

Say you are only interested in the overlaps w/ data3.
The code to filter for would be "..1.".
Say you only want intersect w/ data2 and data3, the code would be ".11.". 

```{r}
category_of_interest3 <- data.frame(
  category = overlap_sizes$category,
  state = str_detect(overlap_sizes$category, 
           "..1.")
) %>% 
  filter(state == T)

combination_matrix3 <- combination_matrix[, category_of_interest3$category]
set_names3 <- set_name(combination_matrix3)
```

The rest is repetition. 

## Quick run down
### Upperleft
```{r}
total_set_sizes3 <- set_size(combination_matrix3) %>% 
  as.data.frame() %>% 
  rename(sizes = ".") %>% 
  mutate(Set = row.names(.)) 

upperleft3 <- total_set_sizes3 %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  ggplot(aes(x = Set, y= sizes)) +
  geom_bar(stat = "identity", aes(fill = Set), alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sizes), 
            size = 5, angle = 90, hjust = 0, y = 1) +
  scale_fill_manual(values = c("orangered3","tomato1", "seagreen", "dodgerblue2"),  # feel free to use some other colors  
                     limits = set_names3) + 
  labs(x = NULL,
       y = "Set size",
       fill = NULL) +
  theme_classic() +
  theme(legend.position = "right",
        text = element_text(size= 14),
        axis.ticks.y = element_blank(),
        axis.text = element_blank()
        ) 

upperleft3
```
### Upperright 
```{r}
upperright3 <- get_legend(upperleft3)
```

### Lowerright
```{r}
overlap_sizes3 <- comb_size(combination_matrix3) %>% 
  as.data.frame() %>% 
  rename(overlap_sizes = ".") %>% 
  mutate(category = row.names(.))

head(overlap_sizes3)
```
```{r}
lowerright3 <- overlap_sizes3 %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>% 
  ggplot(aes(x = category, y = overlap_sizes)) +
  geom_bar(stat = "identity", fill = "grey80", color = NA, alpha = 0.8, width = 0.7) +
  geom_text(aes(label = overlap_sizes, y = 0), 
            size = 5, hjust = 0, vjust = 0.5) +
  labs(y = "Intersect sizes",
       x = NULL) +
  theme_classic() +
  theme(text = element_text(size= 14, color = "black"),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(hjust = 0),
        axis.text = element_blank()
        ) +
  coord_flip()

lowerright3
```

### Lowerleft
```{r}
overlap_matrix3 <- str_split(string = overlap_sizes3$category, pattern = "", simplify = T) %>% 
  as.data.frame() 

colnames(overlap_matrix3) <- set_name(combination_matrix3)

overlap_matrix_tidy3 <- overlap_matrix3 %>% 
  cbind(category = overlap_sizes3$category) %>% 
  pivot_longer(cols = !category, names_to = "Set", values_to = "value") %>% 
  full_join(overlap_sizes3, by = "category") %>% 
  full_join(total_set_sizes3, by = "Set")

head(overlap_matrix_tidy3)
```
```{r}
lowerleft3 <- overlap_matrix_tidy3 %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>%  
  mutate(Set = reorder(Set, sizes)) %>%  
  ggplot(aes(x = Set, y = category))+
  geom_tile(aes(fill = Set, alpha = value), color = "grey30", size = 1) +
  scale_fill_manual(values = c("orangered3","tomato1", "seagreen", "dodgerblue2"), # feel free to use other colors 
                    limits = set_names3) +
  scale_alpha_manual(values = c(0.8, 0),  # color the grid for 1, don't color for 0. 
                     limits = c("1", "0")) +
  labs(x = "Sets", 
       y = "Overlap") +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(color = "black", size= 14),
        panel.grid = element_blank(),
        axis.text = element_blank(),
        )

lowerleft3
```
### Put them together 
```{r}
wrap_plots(upperleft3, upperright3, lowerleft3, lowerright3, 
          nrow = 2, 
          ncol = 2,
          heights = c(1, 2.5), # this time the lower part can be shorter because there're less rows 
          widths = c(0.9, 1)) &
  theme(legend.position = "none")

ggsave("../Results/real_data_upset_3only.svg", height = 4.5, width = 3.5, bg = "white") 
# this time the plot can be shorter 
# the best aspect ratio is when the grids in the colored matrix are perfect squares 
ggsave("../Results/real_data_upset_3only.png", height = 4.5, width = 3.5, bg = "white") 
```

# Extensions
The upset plot can be extended. 
You can add other plots (heatmap, dot plot, boxplot) to the top and side.
These serve as annotation or extra data for the set or intersects. 
With `patchwork` you just need to get creative. 

```{r}
full_data <- data1_full %>%  
  select(location, RPM) %>% 
  full_join(data2_full %>% 
              select(location, RPM), by = "location") %>% 
  full_join(data3_full %>% 
              select(location, RPM), by = "location") %>% 
  full_join(data4_full %>% 
              select(location, RPM), by = "location")

colnames(full_data) <- c("elements", "ovary", "egg", "seedling", "sperm") 
# you just have to change the colnames back to cell types by order

head(full_data)
```


## Extract membership 
You can extract intersect membership using `extract_comb()` from ComplexHeatmap. 
But it's weirdly not vectorized, it only takes one intersect name (e.g., "1000") at a time. 
But we can easily apply it to a list of categories and bind them back to a dataframe. 

```{r}
element_membership_and_value <- overlap_sizes$category %>% 
  map(.f = extract_comb, m = combination_matrix) %>% 
  map(.f = as.data.frame) %>% 
  bind_rows(.id = "id") %>% 
  full_join(
    data.frame(
      id = 1:15,
      category = overlap_sizes$category
    ) %>% 
      mutate(id = as.character(id)), 
    by = "id"
  ) %>% 
  rename(elements = ".x[[i]]")  %>% 
  full_join(full_data, by = "elements")

head(element_membership_and_value)
```

### Extended heatmap for intersects  
Let's say we want another heatmap to the right showing median log10(P) across tissues. 
```{r}
element_membership_and_value_long <- element_membership_and_value %>% 
  pivot_longer(cols = c(ovary, egg, seedling, sperm), names_to = "Set", values_to = "RPM") %>% 
  mutate(log10RPM = log10(RPM)) %>% 
  group_by(category, Set) %>% 
  summarise(median = median(log10RPM)) %>% 
  ungroup()

head(element_membership_and_value_long)
summary(element_membership_and_value_long$median)
quantile(element_membership_and_value_long$median, c(0.025, 0.25, 0.75, 0.9, 0.95, 0.975), na.rm = T)
```
```{r}
extended_heatmap <- element_membership_and_value_long %>% 
  full_join(overlap_sizes, by = "category") %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>% 
  full_join(total_set_sizes, by = "Set") %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  mutate(median_clipped = case_when(
    median > 1 ~ 1, 
    T ~ median
  )) %>% 
  ggplot(aes(x = Set, y = category)) +
  geom_tile(aes(fill = median_clipped), color = "grey30") +
  scale_fill_gradientn(colors = brewer.pal(9, "YlGnBu"), na.value = "white",
                       breaks = c(0.5, 1),
                       labels = c("0.5", "> 1")) +
  labs(x = "Sets",
       y = NULL,
       fill = "median\nlog10(RPM)") +
  theme_minimal() +
  theme(
    text = element_text(size = 14, color = "black"),
    axis.text = element_blank(),
    panel.grid = element_blank(),
    legend.key.width = unit(0.8, "lines")
  )

extended_heatmap
```
```{r}
heatmap_annotation <- total_set_sizes %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  ggplot(aes(x = Set, y = "")) +
  geom_tile(aes(fill = Set), color = "grey80") +
  scale_fill_manual(values = c("orangered3","tomato1", "seagreen", "dodgerblue2"),  # feel free to use some other colors  
                     limits = set_names) + 
  theme_void() +
  theme(legend.position = "none") 


```


## Set metadata 
Let's say data1 and data2 are data derived from control;
data3 and data4 are data derived from a stress treatment. 
We can add that info to the `total_set_sizes` that we made earlier. 
With real data, you can write a metadata table and join that with the `total_set_sizes` table. 

 
### Blank space holder 
```{r}
blank <- data.frame(
  x = 1:10,
  y = 1:10
) %>% 
  ggplot(aes(x = x, y = y)) +
  theme_void()
```

## Put them all together 
```{r}
wrap_plots(upperleft + 
             theme(legend.position = "none"), upperright, wrap_plots(blank, heatmap_annotation, nrow = 2, heights = c(1, 0.3)), 
          lowerleft, lowerright, extended_heatmap, 
          nrow = 2, 
          ncol = 3,
          heights = c(0.3, 1),  
          widths = c(1, 1.3, 1)) # the more stuff on the right, the thinner the left side has to be. 

ggsave("../Results/real_data_upset_extended.svg", height = 4.5, width = 5.5, bg = "white") 
# this time the plot needs to be wider
ggsave("../Results/real_data_upset_extended.png", height = 4.5, width = 5.5, bg = "white") 
```
As you can see the 4-way intersection has the high median expression across all tissues.
The ovary/egg/sperm intersection has very high expression in ovary and egg only, but not in sperm. 

 



