---
title: "Customized_upset_plot"
author: "Chenxin Li"
date: "01/19/2023"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
This is a workflow to generate upset plots with custom aesthetics. 

# Dependencies 
Main packages doing heavy lifting: 

* tidyverse
* ComplexHeatmap
* patchwork 

If you want to save plot as .svg you may need the R package svglite.
If you are using Mac, you may need to install XQuart. [Link](https://www.xquartz.org/)  

```{r}
library(tidyverse) 
library(patchwork) 
library(ComplexHeatmap)

library(RVenn) # Only required if you want Venn diagrams 
library(RColorBrewer) # This is for the colors only, not actually necessary  
library(ggbeeswarm) # Only required if you want quasirandom jittering of dots for mean separation plots 
```


# Data
```{r}
# this chunk generates the data 
# Load your data here when you do your actual thing 
# e.g. data1 <- read_delim(...) 

set.seed(666) # set seed to make simulations reproducible, not needed for actual data 

data1 <- data.frame(
  "elements" = round(runif(n = 30, min = 0, max = 100))
) %>% 
  distinct(elements, .keep_all = T)

data2 <- data.frame(
  "elements" = round(runif(n = 150, min = 0, max = 200))
) %>% 
  distinct(elements, .keep_all = T)

data3 <- data.frame(
  "elements" = round(runif(n = 200, min = 0, max = 300))
) %>% 
  distinct(elements, .keep_all = T)

data4 <- data.frame(
  "elements" = round(runif(n = 60, min = 0, max = 100))
) %>% 
  distinct(elements, .keep_all = T)
```

# Create set list object
```{r}
set_list <- list(data1 = data1$elements, 
                 data2 = data2$elements, 
                 data3 = data3$elements, 
                 data4 = data4$elements) # Change names for your own data 

set_object <- RVenn::Venn(set_list)
```

The order of sets in the list matters. In this case: 

1. data1
2. data2
3. data3
4. data4 

# Venn diagram for 2-3 sets 
```{r}
ggvenn(
  set_object, slice = 1:3, 
  thickness = 0.5,
  alpha = 0.5, 
  fill = brewer.pal(8, "Set2")
) +
  theme_void() +
  theme(
    legend.position = "none"
  )

ggsave("../Results/VennDiagram.svg", height = 4, width = 4, bg = "white")
ggsave("../Results/VennDiagram.png", height = 4, width = 4, bg = "white")
```

# ComplexHeatmap for heavy-lyfting 
```{r}
combination_matrix <- make_comb_mat(set_list)
set_names <- set_name(combination_matrix)
set_names
```

## Total set size 
```{r}
total_set_sizes <- set_size(combination_matrix) %>% 
  as.data.frame() %>% 
  rename(sizes = ".") %>% 
  mutate(Set = row.names(.)) 

upperleft <- total_set_sizes %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  ggplot(aes(x = Set, y= sizes)) +
  geom_bar(stat = "identity", aes(fill = Set), alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sizes), 
            size = 5, angle = 90, hjust = 0, y = 1) +
  scale_fill_manual(values = brewer.pal(4, "Set2"),  # feel free to use some other colors  
                     limits = set_names) + 
  labs(x = NULL,
       y = "Set size",
       fill = NULL) +
  theme_classic() +
  theme(legend.position = "right",
        text = element_text(size= 14),
        axis.ticks.y = element_blank(),
        axis.text = element_blank()
        ) 

upperleft
```
## Legend 
It's not easy to extract legend.
But we can write a function for that.  
```{r}
get_legend <- function(p) {
  tmp <- ggplot_gtable(ggplot_build(p))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  legend
}
```

```{r}
upperright <- get_legend(upperleft)
```


## Overlap sizes
```{r}
overlap_sizes <- comb_size(combination_matrix) %>% 
  as.data.frame() %>% 
  rename(overlap_sizes = ".") %>% 
  mutate(category = row.names(.))

head(overlap_sizes)
```

Categories are ordered by frequency. 
```{r}
lowerright <- overlap_sizes %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>% 
  ggplot(aes(x = category, y = overlap_sizes)) +
  geom_bar(stat = "identity", fill = "grey80", color = NA, alpha = 0.8, width = 0.7) +
  geom_text(aes(label = overlap_sizes, y = 0), 
            size = 5, hjust = 0, vjust = 0.5) +
  labs(y = "Intersect sizes",
       x = NULL) +
  theme_classic() +
  theme(text = element_text(size= 14, color = "black"),
        axis.text =element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(hjust = 0),
        ) +
  coord_flip()

lowerright
```

## Overlap matrix 
```{r}
overlap_matrix <- str_split(string = overlap_sizes$category, pattern = "", simplify = T) %>% 
  as.data.frame() 

colnames(overlap_matrix) <- set_names

overlap_matrix_tidy <- overlap_matrix %>% 
  cbind(category = overlap_sizes$category) %>% 
  pivot_longer(cols = !category, names_to = "Set", values_to = "value") %>% 
  full_join(overlap_sizes, by = "category") %>% 
  full_join(total_set_sizes, by = "Set")

head(overlap_matrix_tidy)
```

```{r}
lowerleft <- overlap_matrix_tidy %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>%  
  mutate(Set = reorder(Set, sizes)) %>%  
  ggplot(aes(x = Set, y = category))+
  geom_tile(aes(fill = Set, alpha = value), color = "grey30", size = 1) +
  scale_fill_manual(values = brewer.pal(4, "Set2"), # feel free to use other colors 
                    limits = set_names) +
  scale_alpha_manual(values = c(0.8, 0),  # color the grid for 1, don't color for 0. 
                     limits = c("1", "0")) +
  labs(x = "Sets",  
       y = "Overlap") +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(color = "black", size= 14),
        panel.grid = element_blank(),
        axis.text = element_blank()
        )

lowerleft
```
 
## Put them all together
```{r}
wrap_plots(upperleft, upperright, lowerleft, lowerright, 
          nrow = 2, 
          ncol = 2,
          heights = c(1, 4), # the more rows in the lower part, the longer it should be
          widths = c(1, 0.8),
          guides = "collect") &
  theme(legend.position = "none")

ggsave("../Results/upset_full.svg", height = 6, width = 3, bg = "white") 
# this should be a tall & skinny plot 
# I prefer .svg, but you can also save as phd or png 
# I will open up the .svg file and mannually adjust the size until it's good
# check that nothing is cut off from the plot 
# png is for twitter posting 
ggsave("../Results/upset_full.png", height = 6, width = 3, bg = "white")
```

# Subsetting the inserctions
## Using ComplexHeatmap 
Filtering of intersections can be done: 
Examples: 

* intersect size >= 4: `m2 <- combination_matrix[comb_size(combination_matrix) >= 4]`.
* degree == 3: `m3 <- combination_matrix[comb_degree(combination_matrix) == 3]`.
* only involving data1 & data2: `m4 <- combination_matrix[c("data1", "data2"), ]`

Say you are only interested in the overlaps w/ data3.
The code to filter for would be "..1.".
Say you only want intersect w/ data2 and data3, the code would be ".11.". 

```{r}
category_of_interest3 <- data.frame(
  category = overlap_sizes$category,
  state = str_detect(overlap_sizes$category, 
           "..1.")
) %>% 
  filter(state == T)

combination_matrix3 <- combination_matrix[, category_of_interest3$category]
set_names3 <- set_name(combination_matrix3)
```

The rest is repetition. 

## Quick run down
### Upperleft
```{r}
total_set_sizes3 <- set_size(combination_matrix3) %>% 
  as.data.frame() %>% 
  rename(sizes = ".") %>% 
  mutate(Set = row.names(.)) 

upperleft3 <- total_set_sizes3 %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  ggplot(aes(x = Set, y= sizes)) +
  geom_bar(stat = "identity", aes(fill = Set), alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sizes), 
            size = 5, angle = 90, hjust = 0, y = 1) +
  scale_fill_manual(values = brewer.pal(4, "Set2"),  # feel free to use some other colors  
                     limits = set_names3) + 
  labs(x = NULL,
       y = "Set size",
       fill = NULL) +
  theme_classic() +
  theme(legend.position = "right",
        text = element_text(size= 14),
        axis.ticks.y = element_blank(),
        axis.text = element_blank()
        ) 

upperleft3
```
### Upperright 
```{r}
upperright3 <- get_legend(upperleft3)
```

### Lowerright
```{r}
overlap_sizes3 <- comb_size(combination_matrix3) %>% 
  as.data.frame() %>% 
  rename(overlap_sizes = ".") %>% 
  mutate(category = row.names(.))

head(overlap_sizes3)
```
```{r}
lowerright3 <- overlap_sizes3 %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>% 
  ggplot(aes(x = category, y = overlap_sizes)) +
  geom_bar(stat = "identity", fill = "grey80", color = NA, alpha = 0.8, width = 0.7) +
  geom_text(aes(label = overlap_sizes, y = 0), 
            size = 5, hjust = 0, vjust = 0.5) +
  labs(y = "Intersect sizes",
       x = NULL) +
  theme_classic() +
  theme(text = element_text(size= 14, color = "black"),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(hjust = 0),
        axis.text = element_blank()
        ) +
  coord_flip()

lowerright3
```

### Lowerleft
```{r}
overlap_matrix3 <- str_split(string = overlap_sizes3$category, pattern = "", simplify = T) %>% 
  as.data.frame() 

colnames(overlap_matrix3) <- set_name(combination_matrix3)

overlap_matrix_tidy3 <- overlap_matrix3 %>% 
  cbind(category = overlap_sizes3$category) %>% 
  pivot_longer(cols = !category, names_to = "Set", values_to = "value") %>% 
  full_join(overlap_sizes3, by = "category") %>% 
  full_join(total_set_sizes3, by = "Set")

head(overlap_matrix_tidy3)
```
```{r}
lowerleft3 <- overlap_matrix_tidy3 %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>%  
  mutate(Set = reorder(Set, sizes)) %>%  
  ggplot(aes(x = Set, y = category))+
  geom_tile(aes(fill = Set, alpha = value), color = "grey30", size = 1) +
  scale_fill_manual(values = brewer.pal(4, "Set2"), # feel free to use other colors 
                    limits = set_names3) +
  scale_alpha_manual(values = c(0.8, 0),  # color the grid for 1, don't color for 0. 
                     limits = c("1", "0")) +
  labs(x = "Sets", 
       y = "Overlap") +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(color = "black", size= 14),
        panel.grid = element_blank(),
        axis.text = element_blank(),
        )

lowerleft3
```
### Put them together 
```{r}
wrap_plots(upperleft3, upperright3, lowerleft3, lowerright3, 
          nrow = 2, 
          ncol = 2,
          heights = c(1, 2.5), # this time the lower part can be shorter because there're less rows 
          widths = c(0.9, 1)) &
  theme(legend.position = "none")

ggsave("../Results/upset_3only.svg", height = 4.5, width = 3.5, bg = "white") 
# this time the plot can be shorter 
# the best aspect ratio is when the grids in the colored matrix are perfect squares 
ggsave("../Results/upset_3only.png", height = 4.5, width = 3.5, bg = "white") 
```

# Extensions
The upset plot can be extended. 
You can add other plots (heatmap, dot plot, boxplot) to the top and side.
These serve as annotation or extra data for the set or intersects. 
With `patchwork` you just need to get creative. 

```{r}
full_data <- rbind(data1, data2, data3, data4) %>% 
  distinct(elements, .keep_all = T) %>% 
  cbind(response = rnorm(mean = 50, sd = 10, n = unite(set_object) %>% length()))
  
# Simulations are for examples only. No need this for actual data 
# Just giving a value to each elements
# Your data will already have this 

head(full_data)
```

## Extract membership 
You can extract intersect membership using `extract_comb()` from ComplexHeatmap. 
But it's weirdly not vectorized, it only takes one intersect name (e.g., "1000") at a time. 
But we can easily apply it to a list of categories and bind them back to a dataframe. 

```{r}
element_membership_and_value <- overlap_sizes$category %>% 
  map(.f = extract_comb, m = combination_matrix) %>% 
  map(.f = as.data.frame) %>% 
  bind_rows(.id = "id") %>% 
  full_join(
    data.frame(
      id = 1:15,
      category = overlap_sizes$category
    ) %>% 
      mutate(id = as.character(id)), 
    by = "id"
  ) %>% 
  rename(elements = ".x[[i]]") %>% 
  full_join(full_data %>% 
              mutate(elements = as.character(elements)), by = "elements")

head(element_membership_and_value)
```

### Mean separation plots for subsets 

```{r}
extended_boxplot <- element_membership_and_value %>% 
  full_join(overlap_sizes, by = "category") %>% 
  mutate(category = reorder(category, -overlap_sizes)) %>% 
  ggplot(aes(x = category, y = response)) +
  geom_quasirandom(shape = 21, color = "white", 
                   fill = "grey20", position = position_jitter(0.1), alpha = 0.8) +
  geom_boxplot(outlier.shape  = NA, width = 0.5, alpha = 0.3, fill = "skyblue1") +
  labs(y = "Response",
       x = NULL) +
  theme_classic() +
  theme(text = element_text(size= 14),
        axis.text=element_text(colour = "black"),
        axis.text.y = element_blank()
        ) +
  coord_flip()

extended_boxplot
```
In your actual data you can color the boxes or dots w/ factors of interest.  
Note: box plots aren't that useful for categories with n < 30.
This section can easily be replaced by a heatmap or a dot plot showing the mean of each subset. 

## Set metadata 
Let's say data1 and data2 are data derived from control;
data3 and data4 are data derived from a stress treatment. 
We can add that info to the `total_set_sizes` that we made earlier. 
With real data, you can write a metadata table and join that with the `total_set_sizes` table. 

```{r}
metadata <- data.frame(
  Set = c("data1", "data2", "data3", "data4"),
  Condition = c("ctrl", "ctrl", "stress", "stress")
)

total_set_sizes_metadata <- total_set_sizes %>% 
  full_join(metadata, by = "Set") 

total_set_sizes_metadata
```

### Metadata annotation by heatmap 
```{r}
metadata_heatmap <- total_set_sizes_metadata %>% 
  mutate(Set = reorder(Set, sizes)) %>% 
  ggplot(aes(x = Set, y = "")) +
  geom_tile(aes(fill = Condition), alpha = 0.8) +
  scale_fill_manual(values = c("grey80", "tomato1"),  # feel free to use some other colors  
                     limits = c("ctrl", "stress")) + 
  guides(fill = guide_legend(title.position = "top")) +
  labs(x = NULL,
       y = NULL,
       fill = "Condition") +
  theme_void() +
  theme(legend.position = "bottom",
        text = element_text(size= 14),
        axis.ticks = element_blank(),
        axis.text = element_blank()
        ) 

metadata_heatmap
```
### Blank space holder 
```{r}
blank <- data.frame(
  x = 1:10,
  y = 1:10
) %>% 
  ggplot(aes(x = x, y = y)) +
  theme_void()
```

## Put them all together 
```{r}
wrap_plots(upperleft + 
             theme(legend.position = "none"), upperright, blank, 
          lowerleft, lowerright, extended_boxplot, 
          metadata_heatmap, blank, blank,
          nrow = 3, 
          ncol = 3,
          heights = c(0.3, 1, 0.05),  
          widths = c(0.7,0.8, 1)) # the more stuff on the right, the thinner the left side has to be. 

ggsave("../Results/upset_extended.svg", height = 5.8, width = 4.7, bg = "white") 
# this time the plot needs to be wider
ggsave("../Results/upset_extended.png", height = 5.8, width = 4.7, bg = "white") 
```

That's pretty much it. 



